Time to stop; agent-based modelling of chemoaffinity with competition

### 2022-09-02 paper/main.tex
Friday WIP

%%%%% Describing the simulation
We simulated the retina and tectum as unit square regions and defined retinal receptor expression and tectal ligand expression across these squares (Fig.\,\ref{f:ex}B-D).
The simulated retina possessed $n$=400 (20$\times$20) retinal ganglion cells; the tectum was similarly arranged as 20$\times$20 discrete, square elements, with defined levels of expression for each ligand.
One spatial unit corresponds to 2~mm on the retina or tectum for mouse \citep{reber_relative_2004}.

Each retinal ganglion cell (RGC) projects $M=4$ growth cones (referred to as \emph{branches}) which carry a set of $N$ receptors, $r_i$, indexed by $i$ and expressed at levels determined by the cell soma's location on the retinal grid.
($N$ ligands, $l_i$, are also expressed by RGCs, but these are not used for chemotaxis.)
The expression of each receptor varies only with respect to one spatial dimension.
%
We defined 4 receptor expression gradients arranged in orthogonal pairs with the gradient of $r_0$ being orthogonal to that of $r_1$. $r_2$, whose gradient is opposite to $r_0$, was orthogonal to $r_3$ (Fig.\,\ref{f:ex}B).
%
Because there is convincing evidence that EphA and EphB receptors are expressed in exponentially increasing patterns \citep{reber_relative_2004,feldheim_genetic_2000,brown_topographic_2000,koulakov_stochastic_2004}, we use an exponential form for retinal receptor expressions, in common with other modelling studies~\citep{reber_relative_2004,koulakov_stochastic_2004,simpson_simple_2011}.
We adopt the same precise form for the retinal receptor expression as \citet{simpson_simple_2011}, $f(x,y) = 1.05 + 0.26 \exp(2.3 u)$, where $u$ gives the direction with which the expression increases and is substituted by $(1-x)$ for $r_0$, $(1-y)$ for $r_1$, $x$ for $r_2$ and by $y$ for $r_3$ (Fig.\,\ref{f:ex}B).

Cells on the tectum express ligands, $L_i$, for the retinal receptors, also in orthogonal pairs of gradients.
Although several studies model tectal ligand expression with exponential functions \citep{koulakov_stochastic_2004}, the experimental evidence for the form of tectal ligand expression is more ambiguous than that for retinal receptor expression (Fig.\,\ref{f:expressions} summarises results from the literature).
We kept in mind the possibility that tectal ligand expression may be modelled by another function, but initially, we set it to the same exponential used for retinal expression and used the substitutions $y$, $x$, $(1-y)$ or $(1-x)$ in place of $u$ to obtain the ligand expression functions for $L_0$, $L_1$, $L_2$ and $L_3$ (Fig.\,\ref{f:ex}D).

@@

Fig.\,\ref{f:GJ} shows the simulation results with these parameters after 1500 time steps.
Fig.\,\ref{f:GJ}A indicates the colour code used for retinal cell position; dorsal retinal cells are indicated by green, nasal cells by red. Lightness gives position along the relevant axis so that cells at the dorsal-nasal corner are yellow and those at the ventral-temporal corner are black.
Fig.\,\ref{f:GJ}E shows the expected layout of retinal ganglion cell axons on the tectum.
The axons retain the colour of their originating cell on the retina.
Nasal cell axons should be arranged caudally; dorsal axons laterally; axons should be evenly spaced and the pattern should span the entire tectum.
Fig.\,\ref{f:GJ}, panels B--D show `fish net' plots of the average location of the 4 branches of each axon (the axon's centroid) at three timepoints.
Fig.\,\ref{f:GJ}B indicates the randomised placement of axon branches at the rostral end of the tectum as specified by the initial conditions (Eq.\,\ref{e:ic}).
In these plots, lines are drawn between axons that are expected to arrange themselves adjacently to one another by the end of the simulation.
The number of line crossings in this plot gives an indication of the disorder in the arrangement.
Fig.\,\ref{f:GJ}G shows the positions of the individual branches (i.e.~growth cones) at $t=1500$ that contribute to the plot in Fig.\,\ref{f:GJ}D.
By this time, the topological order matches that shown in Fig.\,\ref{f:GJ}E.
Opposing gradients have sorted the axons along the two axes to form a grid of cells.
The arrangement spans the full tectum due to competition between the growth cones (compare with the chemotaxis-only model in Fig.\,\ref{f:G}).
Fig.\,\ref{f:GJ}F shows the two pattern metrics; $\epsilon$ (distance error) and $\eta$ (number of crossings).
For the wildtype pattern, $\eta_t=0$, as in Fig.\,\ref{f:GJ}E.
Fig.\,\ref{f:GJ}H shows selected axons (one from the centre and one from each corner of the retina) and their position history.

@@

After finding an optimum set of parameters for the wildtype arrangement, we tested our model with a number of simulations of surgical manipulations (for reviews, see \citet{udin_formation_1988} and \citet{goodhill_retinotectal_1999}).
Figs.~\ref{f:GJsurg}A and \ref{f:GJsurg}B show the results of tectal graft rotation manipulations in which a square patch of tectal tissue is cut out and rotated $90\degree$ or $180\degree$ \citep{chung_observations_1978}.
%
To simulate the surgery, we rotated an $8\times8$ square of the tectum shown in Fig\,\ref{f:ex}D, carrying the ligand expressions with the rotation (resulting in Fig.\,\ref{f:trot90}, left column).
After the manipulation, we re-computed the tectal ligand expression gradients. Fig.\,\ref{f:trot90} demonstrates that the most significant change in the gradient is around the border of the graft, due to discontinuities in the ligand expression.
The manipulated gradients, interacting with unchanged axon branches, lead to a qualitative rotation of axons that were destined for the graft area, indicated by a 90$\degree$ or 180$\degree$ `twist' of the axon centroid colours.
Long connecting lines in the fish net show that a number of axons from near the edges of the tectum have been drawn into the graft region.
The pattern metric, $\epsilon$, is shown in the right hand graphs of Fig.\,\ref{f:GJsurg}A \& B.
We plot $\epsilon$ (the distance error) for the entire map (blue) along with the error for the graft region (gold) and the surround region (red).
The overall distance error for the rotation manipulations (0.182 or 0.2752)  is greater than for the wildtype system (0.04).
The pattern within the graft exhibits greater average distance error than for the surrounding tissue.
The system does not achieve a perfect reproduction of the target layout suggested by the experimental result (which is shown in the inset) but it is in qualitative agreement.
%
Fig.\,\ref{f:GJsurg}C shows the result of a similar manipulation in which two grafts are cut out of the tectum and swapped.
There is considerable disorder in the simulated pattern of axon centroids but a line of red/yellow agents that would been located caudally has advanced rostrally and a number of green agents are more caudal than in the wildtype arrangement, consistent with experimental observations.
% Note: the graph shows the pattern metric for a region that covers BOTH grafts and for the region that is not the two grafts.

Fig.\,\ref{f:GJsurg}D shows the result of a retinal ablation, in which the temporal half of the retina is surgically removed and the remaining RGC axons are allowed to develop on the unmanipulated tectum.
According to a pure chemoaffinity theory, the surviving retinal axons should retain their chemical labels and they would be expected to find their usual locations on the tectum, resulting in a tectum which is only half populated with axons.
The experimental observation is that the axons arrange themselves across the whole tectum, retaining their topographic order (inset, \citet{attardi_preferential_1963,schmidt_expansion_1978}).
The fish-net plot of the simulation indicates that the competition in the system successfully `stretches out' the surviving axons to achieve a close representation of the experimentally observed result.
In this figure, the right-hand graph shows the simulated position of each axon centroid on the rostro-caudal tectal axis plotted against its position on the nasal-temporal retinal axis in red triangles, with the experimental positions plotted in blue squares.
%
Fig.\,\ref{f:GJsurg}E shows the result of a tectal ablation. Here the retina is left un-modified, but the caudal half of the tectum is surgically removed. Experimental observation \citep{yoon_reorganization_1971,sharma_reformation_1972} shows that RGC axons `crush up' on the remaining rostral section, retaining topographic order and creating a complete map on the available tectal tissue. Our model clearly replicates this result, indicated by both fish-net and tectal/retinal axes graphs, because the chemotaxis component of the model is based on balanced, relative interactions between opposing gradients, rather than absolute labels or expression measurements.
%
A related ablation experiment is the so-called `mismatch', in which the temporal half of the retina is ablated along with the caudal part of the tectum to which it normally maps, again challenging the pure chemoaffinity model \citep{horder_retention_1971}. Our model replicates this manipulation (Fig.\,\ref{f:GJsurg}F) for the same reason as in the simple tectal ablation of Fig.\,\ref{f:GJsurg}E; opposing gradients remain on the surviving tectum allowing RGC axons to find their relative ordering.
%
The final surgical manipulation which we investigated was the `compound eye' experiment, in which two nasal retina halves are fused together \citep{gaze_retino-tectal_1963,fawcett_retinotectal_1982}. Within the compound retina there exist matched pairs of retinal ganglion cells, one from each nasal half, whose receptor and ligand expressions cannot be distinguished. These form overlaid maps, each of which spans the entire tectum (Fig.\,\ref{f:GJsurg}G).

@@

In the early 2000s, a series of genetic experiments added to the set of manipulations that a model of the retinotectal system would need to
explain.
First, \citet{brown_topographic_2000} showed that a selective RGC knock-in of EphA3 receptors led to an interesting modulation of the maps found in the tectum.
For the strongest knock-in (a homozygote manipulation) two maps were formed, with knock-in cells finding termination zones more rostral than their wildtype retinal neighbours.
By enhancing the quantity of EphA receptors expressed by half of all retino-tectal axons, they demonstrated the importance of retinal receptor expression for the arrangement of the tectal map.
In the weaker, heterozygote manipulation, the maps were still separated, but the separation `collapsed' towards the rostral end of the tectum (see \citet{brown_topographic_2000}, Fig\,5).
This led to a second study showing that with an additional knock-down of EphA4 receptors in \emph{all} RGCs, the retinal origin of the collapse point was shifted temporally and all heterozygote EphA3 knock-in cells were pushed even further rostrally (\citet{reber_relative_2004}, their Fig.\,3e).
We tested whether our model would reproduce these results.

An assumption made in these papers is that all EphA receptor subtypes contribute to a summed EphA expression level, $\sum\mathrm{EphA}$, which increases from a low level at the nasal side of the retina, to a high level at the temporal side.
The precise contribution of each subtype to the overall level is assumed not to be important.
In our model, receptor $r_0$ increases in the retina in a nasal to temporal direction (Fig\,\ref{f:ex}B) and so we manipulated $r_0$ to recreate the Brown et al.~and Reber et al.~experiments.
In half of the RGCs, we knocked EphA3 in by adding to the value of $r_0$ in these cells either 1.0 (heterozygote EphA3 ki/+) or 4.0 (homozygote EphA3 ki/ki).
To include the additional knock-down of EphA4 from \citet{reber_relative_2004}, we subtracted 0.7 from $r_0$ in \emph{all} RGCs.
Fig.\,\ref{f:GJeph}A shows, for the single EphA3 knock-in (EphA3 ki/+), branch positions, axon centroids and final rostral-caudal position on the tectum plotted versus the originating nasal-temporal retinal coordinate for time step 1500.
The left panel shows branches coloured in the red/green duochrome scheme according to their origin location on the retina.
In the middle panel, axon centroids of RGCs which were treated with the knock-in manipulation are coloured red, wildtype cells are blue.
The map of treated cells on the tectum was shifted slightly rostrally, and the gap between the red and blue curves became smaller for more temporal RGCs. However, it can't be described as a map collapse (the term used in \citet{brown_topographic_2000}) because a bootstrapped studentized t-test of the two sets of points indicates that they are drawn from statistically different distributions (p<0.0025).
% While the convergence of the manipulated and wildtype lines was less abrupt than in the experimental result of \citet{brown_topographic_2000}, it was comparable with the result from \citet{simpson_simple_2011}.
Fig.\,\ref{f:GJeph}B shows the same trio of graphs for the homozygote double knock-in, EphA3 ki/ki. Here, the treated axons were pushed further towards the rostral end of the tectum and the untreated axons were pushed caudally.
The maps were completely separated; as in the experiment, there was no collapse point.

The rostral movement of treated axons occurred due to their increased expression of EphA ($r_0$), which increased the simulated interaction with the tectal rostral-caudal ligand gradient.
The opposing receptor expression ($r_2$) was unchanged by the manipulation and so those axons with increased $r_0$ expression were pulled to a more rostral position than they would otherwise have taken.
The caudal movement of wildtype axons, whose receptor expression in Figs.\,\ref{f:GJeph}A-B was unchanged, occurred due to competition with the most caudal of the rostrally-drawn treated axons.

Figs.\,\ref{f:GJeph}C show the results for the \citet{reber_relative_2004} manipulation, in which the EphA expression was selectively enhanced (by EphA3 knock-in) in 50\% of cells and knocked down (by EphA4 knock-down) in all cells.
As in Fig.\,\ref{f:GJeph}A, the red and blue lines converge, but there is no map collapse (p<0.008).
The significant rostral movement of the treated axons seen in \citet{reber_relative_2004}, where not a single treated axon was found more than 30\% along the rostral-caudal axis, was absent from our result.
In fact, the knock-in/knock-down cells find locations slightly more \emph{caudal} than those in Fig.\,\ref{f:GJeph}A because while $r_0$ is increased by 1 from the knock-in, the knock-down of 0.7 reduces these cells' interaction with tectal ephrin-A ($L_0$) compared with those in the EphA3 knock-in/EphA4 wildtype simulation.
It is important to note that these results emerge purely from modifications to the receptor expression levels; the model is otherwise unchanged.
The lack of collapse points or the more rostral positions of knock-in/knock-down cells suggests some interaction effect between EphA4 and other EphA receptors that is not captured by our model.
We address this in the next section.

\subsection*{Two new mechanisms to address genetic results: Cluster size and $r_2$ collapse}

The previous approach assumed that all EphA receptor subtypes (EphA3, EphA4, EphA5, etc) transmit the same signal to induce indistinguishable function.
This assumption is difficult to support, given that the effect caused by knock-\emph{in} of EphA3 (a shift of nasal RGC axons towards the rostral tectum) is enhanced by the knock-\emph{down} of EphA4.
We therefore changed our assumption to the following: Most EphA receptors share the same function (EphA3, EphA5 etc) but EphA4 is different.
We redefined $r_{0,b}$ as the expression of EphA receptors (excluding EphA4) on branch $b$ (Fig.\,\ref{f:clustermech}A), and introduced $r_{\!\scriptscriptstyle A4,b}$ as the expression of free EphA4 receptors (Fig.\,\ref{f:clustermech}B). Here, we assume that cis-interactions between retinal ephrinA ligands and retinal EphA4 receptors lead to bound and un-bound EphA4 receptors \citep{hornberger_modulation_1999}. The number of cis-bound EphA4 receptors is taken as $r_{A4}^{\mathrm{cis}} = 0.15\,l_0$ and $r_{\!\scriptscriptstyle A4, b} = 3.5 - r_{A4}^{\mathrm{cis}}$ where the value 0.15 is a binding weight and 3.5 represents the spatially homogeneous retinal   expression of EphA4 (Fig.\,\ref{f:clustermech}C).

It is known that EphA receptors form clusters whose presence is necessary for ligand attachments to induce signal transmission and that cluster size affects signal size \citep{nikolov_ephephrin_2013}. It is also known that that some receptor sub-types may compete for ligand attachment \citep{fiore_regulation_2019}.
Additionally, EphA4 receptors make `side attachments' to clusters of EphA receptors, possibly regulating the cluster size \citep{nikolov_ephephrin_2013}.
%
We focused on the idea that signal strength is related to receptor cluster size and defined $c_i$, the effective cluster size for receptor $i$.
We set $c_i=1$ for $i=1$, $2$ or $3$ and investigated a simple relationship between EphA cluster size, $c_0$, and $r_{\!\scriptscriptstyle A4,b}$, the density of EphA4 receptors on branch $b$:
\begin{equation}
    c_0 = 1/r_{\!\scriptscriptstyle A4,b}
\end{equation}
%
Eq.\,\ref{e:G} was modified to include the cluster size:
%
\begin{equation}\label{e:Gcs}
\mathbf{G} = m_{\!_G}\,\sum_i^N F_i\,c_i\,r_{i,b} \nabla L_i(\mathbf{x})
\end{equation}

$c_i\,r_{i,b}$ is the effective signal strength, $s_i$, for receptor $i$ and differs from $r_{i,b}$ only for $i=0$ for which (dropping the branch subscript) $s_0 = r_0/r_{\!\scriptscriptstyle A4}$.
Under this relationship, the signal strength changes as $r_0$ increases (simulating the knock-in of EphA3) and also as $r_{\!\scriptscriptstyle A4}$ decreases (simulating EphA4 knock-down).
%(This is reminiscent of the divisive relationship suggested by the work of Brown, Reber and co-workers who showed that when the ratio of EphA in treated to EphA in untreated cells drops below a threshold, the mapping system directs the cells to the same tectal location.)
Fig\,\ref{f:clustermech}C shows $s_0$ under knock-in, knock-down and combined knock-in/knock-down conditions.
Fig\,\ref{f:clustermech}D shows the logic that is assumed to be enacted by the EphA/EphA4 and $r_2$ receptors to implement the $r_2$ collapse mechanism. We assume that there is a coupling between $r_0$, $r_{A4}$ expression and $r_2$ expression. In a region where $r_0$ exceeds some threshold, $h_0$, and $r_{A4}$ is \emph{below} another threshold, $h_1$, the effectiveness of $r_2$ is reduced to one fifth of its normal value.

>>>>

%%%%% Describing the simulation
We simulated the retina and tectum as unit square regions and defined retinal receptor expression and tectal ligand expression across these squares in the directions defined in Fig.\,\ref{f:ex}. 
The simulated retina possessed $n$=400 (20$\times$20) retinal ganglion cells; the tectum was similarly arranged as 20$\times$20 discrete, square elements, with defined levels of expression for each ligand. 
One spatial unit corresponds to 2~mm on the retina or tectum for mouse \citep{reber_relative_2004}.

Each retinal ganglion cell (RGC) projects $M=4$ growth cones (referred to as \emph{branches}) which carry a set of $N$ receptors, $r_i$, indexed by $i$ and expressed at levels determined by the cell soma's location on the retinal grid. 
($N$ ligands, $l_i$, are also expressed by RGCs, but these are not used by this chemotaxis model.) 
The expression of each receptor varies only with respect to one spatial dimension.
%
We defined 4 receptor expression gradients arranged in orthogonal pairs with the gradient of $r_0$ being orthogonal to that of $r_1$. $r_2$, whose gradient is opposite to $r_0$, was orthogonal to $r_3$ (Fig.\,\ref{f:ex}B).
%
Because there is convincing evidence that EphA and EphB receptors are expressed in exponentially increasing patterns \citep{reber_relative_2004,feldheim_genetic_2000,brown_topographic_2000,koulakov_stochastic_2004}, we use an exponential form for retinal receptor expressions, in common with other modelling studies~\citep{reber_relative_2004,koulakov_stochastic_2004,simpson_simple_2011}.
We adopt the same precise form for the retinal receptor expression as \citet{simpson_simple_2011}, $f(x,y) = 1.05 + 0.26 \exp(2.3 u)$, where $u$ gives the direction with which the expression increases and is substituted by $(1-x)$ for $r_0$, $(1-y)$ for $r_1$, $x$ for $r_2$ and by $y$ for $r_3$ (Fig.\,\ref{f:ex}).

Cells on the tectum express ligands, $L_i$, for the retinal receptors, also in orthogonal pairs of gradients.
Although several studies model tectal ligand expression with exponential functions \citep{koulakov_stochastic_2004}, the experimental evidence for the form of tectal ligand expression is more ambiguous than that for retinal receptor expression (Fig.\,\ref{f:ex} summarises results from the literature). 
We kept in mind the possibility that tectal ligand expression may be modelled by another function, but initially, we set it to the same exponential used for retinal expression and used the substitutions $y$, $x$, $(1-y)$ or $(1-x)$ in place of $u$ to obtain the ligand expression functions for $L_0$, $L_1$, $L_2$ and $L_3$ (Fig.\,\ref{f:ex}).

@@

Fig.\,\ref{f:GJ} shows the simulation results with these parameters after 1500 time steps. 
Fig.\,\ref{f:GJ}A indicates the colour code used for retinal cell position; dorsal retinal cells are indicated by green, nasal cells by red. Lightness gives position along the relevant axis so that cells at the dorsal-nasal corner are yellow and those at the ventral-temporal corner are black.
Fig.\,\ref{f:GJ}E shows the expected layout of retinal ganglion cell axons on the tectum.
The axons retain the colour of their originating cell on the retina.
Nasal cell axons should be arranged caudally; dorsal axons laterally; axons should be evenly spaced and the pattern should span the entire tectum.
Fig.\,\ref{f:GJ}, panels B--D show `fish net' plots of the average location of the 4 branches of each axon (the axon's centroid) at three timepoints. 
Fig.\,\ref{f:GJ}B indicates the randomised placement of axon branches at the rostral end of the tectum as specified by the initial conditions (Eq.\,\ref{e:ic}).
In these plots, lines are drawn between axons that are expected to arrange themselves adjacently to one another by the end of the simulation. 
The number of line crossings in this plot gives an indication of the disorder in the arrangement.
Fig.\,\ref{f:GJ}G shows the positions of the individual branches (i.e.~growth cones) at $t=1500$ that contribute to the plot in Fig.\,\ref{f:GJ}D.
By this time, the topological order matches that shown in Fig.\,\ref{f:GJ}E.
Opposing gradients have sorted the axons along the two axes to form a grid of cells.
The arrangement spans the full tectum due to competition between the growth cones (in the chemotaxis-only model with $m_{\!_X}=0$, axons cover only a fraction of the tectal tissue).
Fig.\,\ref{f:GJ}F shows the two pattern metrics; $\epsilon$ (distance error) and $\eta$ (number of crossings).
For the wildtype pattern, $\eta_t=0$, as in Fig.\,\ref{f:GJ}E. 
Fig.\,\ref{f:GJ}H shows selected axons (one from the centre and one from each corner of the retina) and their position history.

@@

After finding an optimum set of parameters for the wildtype arrangement, we tested our model with a number of simulations of surgical manipulations (for reviews, see \citet{udin_formation_1988} and \citet{goodhill_retinotectal_1999}). 
Figs.~\ref{f:GJsurg_grafts}A and \ref{f:GJsurg_grafts}B show the results of tectal graft rotation manipulations in which a square patch of tectal tissue is cut out and rotated $90\degree$ or $180\degree$ \citep{chung_observations_1978}. 
%
To simulate the surgery, we rotated an $8\times8$ square of the tectum shown in Fig\,\ref{f:ex}D, carrying the ligand expressions with the rotation (resulting in Fig.\,\ref{f:trot90}, left column). 
After the manipulation, we re-computed the tectal ligand expression gradients. Fig.\,\ref{f:trot90} demonstrates that the most significant change in the gradient is around the border of the graft, due to discontinuities in the ligand expression. 
The manipulated gradients, interacting with unchanged axon branches, lead to a qualitative rotation of axons that were destined for the graft area, indicated by a 90$\degree$ or 180$\degree$ `twist' of the axon centroid colours. 
Long connecting lines in the fish net show that a number of axons from near the edges of the tectum have been drawn into the graft region. 
The pattern metric, $\epsilon$, is shown in the right hand graphs of Fig.\,\ref{f:GJsurg_grafts}A \& B. 
We plot $\epsilon$ (the distance error) for the entire map (blue) along with the error for the graft region (gold) and the surround region (red). 
The overall distance error for the rotation manipulations (0.182 or 0.2752)  is greater than for the wildtype system (0.04). 
The pattern within the graft exhibits greater average distance error than for the surrounding tissue.
The system does not achieve a perfect reproduction of the target layout suggested by the experimental result (which is shown in the inset) but it is in qualitative agreement.
%
Fig.\,\ref{f:GJsurg_grafts}C shows the result of a similar manipulation in which two grafts are cut out of the tectum and swapped. 
There is considerable disorder in the simulated pattern of axon centroids but a line of red/yellow agents that would been located caudally has advanced rostrally and a number of green agents are more caudal than in the wildtype arrangement, consistent with experimental observations.
% Note: the graph shows the pattern metric for a region that covers BOTH grafts and for the region that is not the two grafts.

Fig.\,\ref{f:GJsurg_ablate}A shows the result of a retinal ablation, in which the temporal half of the retina is surgically removed and the remaining RGC axons are allowed to develop on the unmanipulated tectum.
According to a pure chemoaffinity theory, the surviving retinal axons should retain their chemical labels and they would be expected to find their usual locations on the tectum, resulting in a tectum which is only half populated with axons. 
The experimental observation is that the axons arrange themselves across the whole tectum, retaining their topographic order (inset, \citet{attardi_preferential_1963,schmidt_expansion_1978}). 
The fish-net plot of the simulation indicates that the competition in the system successfully `stretches out' the surviving axons to achieve a close representation of the experimentally observed result. 
In this figure, the right-hand graph shows the simulated position of each axon centroid on the rostro-caudal tectal axis plotted against its position on the nasal-temporal retinal axis in red triangles, with the experimental positions plotted in blue squares.
%
Fig.\,\ref{f:GJsurg_ablate}B shows the result of a tectal ablation. Here the retina is left un-modified, but the caudal half of the tectum is surgically removed. Experimental observation \citep{yoon_reorganization_1971,sharma_reformation_1972} shows that RGC axons `crush up' on the remaining rostral section, retaining topographic order and creating a complete map on the available tectal tissue. Our model clearly replicates this result, indicated by both fish-net and tectal/retinal axes graphs, because the chemotaxis component of the model is based on balanced, relative interactions between opposing gradients, rather than absolute labels or expression measurements.
%
A related ablation experiment is the so-called `mismatch', in which the temporal half of the retina is ablated along with the caudal part of the tectum to which it normally maps, again challenging the pure chemoaffinity model \citep{horder_retention_1971}. Our model replicates this manipulation (Fig.\,\ref{f:GJsurg_ablate}C) for the same reason as in the simple tectal ablation of Fig.\,\ref{f:GJsurg_ablate}B; opposing gradients remain on the surviving tectum allowing RGC axons to find their relative ordering.
%
The final surgical manipulation which we investigated was the `compound eye' experiment, in which two nasal retina halves are fused together \citep{gaze_retino-tectal_1963,fawcett_retinotectal_1982}. Within the compound retina there exist matched pairs of retinal ganglion cells, one from each nasal half, whose receptor and ligand expressions cannot be distinguished. These form overlaid maps, each of which spans the entire tectum (Fig.\,\ref{f:GJsurg_ablate}D).

@@

In the early 2000s, a series of genetic experiments added to the set of manipulations that a model of the retinotectal system would need to
explain. 
First, \citet{brown_topographic_2000} showed that a selective RGC knock-in of EphA3 receptors led to an interesting modulation of the maps found in the tectum. 
For the strongest knock-in (a homozygote manipulation) two maps were formed, with knock-in cells finding termination zones more rostral than their wildtype retinal neighbours. 
By enhancing the quantity of EphA receptors expressed by half of all retino-tectal axons, they demonstrated the importance of retinal receptor expression for the arrangement of the tectal map. 
In the weaker, heterozygote manipulation, the maps were still separated, but the separation `collapsed' towards the rostral end of the tectum (see \citet{brown_topographic_2000}, Fig\,5). 
This led to a second study showing that with an additional knock-down of EphA4 receptors in \emph{all} RGCs, the retinal origin of the collapse point was shifted temporally and all heterozygote EphA3 knock-in cells were pushed even further rostrally (\citet{reber_relative_2004}, their Fig.\,3e). 
For our model to successfully reproduce these results, it should therefore: i)~Demonstrate a complete separation of maps for a `large' knock-in of EphA3; ii)~Show a separation of maps for a smaller knock-in of EphA3 that is greater for axons of nasal origin. iii)~For the smaller knock-in, a collapse point at 76\% of the N--T axis; iv)~Show a considerable rostral shift of cells for which EphA3 is knocked in and EphA4 is knocked down; and v)~Show a collapse point for the knock-in/knock-down cells and knock-down only cells at around 88\% of the N--T axis.

An assumption made in these papers is that all EphA receptor subtypes contribute to a summed EphA expression level, $\sum\mathrm{EphA}$, which increases from a low level at the nasal side of the retina, to a high level at the temporal side. 
The precise contribution of each subtype to the overall level is assumed not to be important. 
In our model, receptor $r_0$ increases in the retina in a nasal to temporal direction (Fig\,\ref{f:ex}B) and so we treated $r_0$ as being equivalent to $\sum\mathrm{EphA}$ and manipulated it to attempt a recreation of the \citet{brown_topographic_2000}~and \citet{reber_relative_2004}~experiments. 
In half of the RGCs, we knocked EphA3 in by adding to the value of $r_0$ in these cells either 0.8 (heterozygote EphA3 ki/+) or 3.2 (homozygote EphA3 ki/ki). 
To include the additional heterozygote knock-down of EphA4 (EphA4 +/-) from \citet{reber_relative_2004}, we subtracted 1.2 from $r_0$ in \emph{all} RGCs.
Fig.\,\ref{f:GJeph}A shows, for the single EphA3 knock-in (EphA3 ki/+), branch positions, a plot of axon centroids and final rostral-caudal position on the tectum versus the originating nasal-temporal retinal coordinate for time step 1500. 
The left panel shows branches coloured in the red/green duochrome scheme according to their origin location on the retina.
In the middle panel, axon centroids of RGCs which were treated with the knock-in manipulation are coloured red, wildtype cells are blue. 
The right hand panel mapping final R--C position versus N--T origin  (and its inset which shows the experimental result) follow the same red/blue colour scheme.
The map of treated cells (red) was shifted slightly rostrally, and the gap between the red and blue curves became smaller for more temporal RGCs. However, it can't be described as a map collapse (the term used in \citet{brown_topographic_2000}) because the curves remain separated for all N--T values. Regardless of the value of the ratio between treated and untreated cells, treated temporal axons are drawn to a significantly more rostral position than those without the knock-in. A bootstrapped studentized t-test of the two sets of points indicates a low probability, $p<10^{-4}$, that they are drawn from statistically identical distributions.
% While the convergence of the manipulated and wildtype lines was less abrupt than in the experimental result of \citet{brown_topographic_2000}, it was comparable with the result from \citet{simpson_simple_2011}.

Fig.\,\ref{f:GJeph}B shows the same trio of graphs for the homozygote double knock-in, EphA3 ki/ki. Here, the treated axons were pushed further towards the rostral end of the tectum and the untreated axons were pushed caudally due to competition with the treated cells. 
The maps were completely separated. As in the experiment, there was no collapse point.
The rostral movement of treated axons occurred due to their increased expression of EphA ($r_0$), which increased the simulated interaction with the tectal rostral-caudal ligand gradient. 
The opposing receptor expression ($r_2$) was unchanged by the manipulation and so those axons with increased $r_0$ expression were pulled to a more rostral position than they would otherwise have taken. 
The caudal movement of wildtype axons, whose receptor expression in Figs.\,\ref{f:GJeph}A-B was unchanged, occurred due to competition with the most caudal of the rostrally-drawn treated axons.
Overall, the correspondence between the experimental result and the simulated homozygous knock-in is close.

Figs.\,\ref{f:GJeph}C show the results for the \citet{reber_relative_2004} manipulation, in which the EphA expression was selectively enhanced (by EphA3 knock-in) in 50\% of cells and knocked down (by EphA4 knock-down) in all cells.
As in Fig.\,\ref{f:GJeph}A, the red and blue lines are separated, but there is no map collapse ($p<10^{-4}$). 
The significant rostral movement of the treated axons seen in \citet{reber_relative_2004}, where not a single treated axon was found more than 30\% along the rostral-caudal axis, was absent from this result. 
In fact, the knock-in/knock-down cells find locations slightly more \emph{caudal} than those in Fig.\,\ref{f:GJeph}A because while $r_0$ is increased by 0.8 from the knock-in, the knock-down of 1.2 reduces these cells' interaction with tectal ephrinA ($L_0$) compared with those in the EphA3 knock-in/EphA4 wildtype simulation.

It is important to note that these results emerge purely from modifications to the receptor expression levels; the model is otherwise unchanged. 
The lack of collapse points or the more rostral positions of knock-in/knock-down cells suggests that there is an interaction effect between EphA4 and other EphA receptors that has not been captured by this model.
We address possible interaction effects in the next section.

\subsection*{Two mechanisms to address the genetic results: Cluster size and $r_2$ collapse}

The previous approach assumed that all EphA receptor subtypes (EphA3, EphA4, EphA5, etc) transmit the same signal to induce indistinguishable function.
This assumption is difficult to support, given that the effect caused by knock-\emph{in} of EphA3 (a shift of nasal RGC axons towards the rostral tectum) is enhanced by the knock-\emph{down} of EphA4. 
We therefore changed our assumption to the following: Most EphA receptors share the same function (EphA3, EphA5 etc) but EphA4 is different. 
We redefined $r_{\!\scriptscriptstyle 0,b}$ as the expression of EphA receptors (excluding EphA4) on branch $b$ (Fig.\,\ref{f:clustermech}A), and introduced $r_{\!\scriptscriptstyle A4,b}$ as the expression of EphA4 receptors (Fig.\,\ref{f:clustermech}B). EphA4 receptors are expressed in a spatially inhomogeneous pattern~\citep{reber_relative_2004} for which we choose the value $r_{\!\scriptscriptstyle A4} = 3.5$, dropping the branch subscript.  We assume that cis-interactions between retinal ephrinA ligands and retinal EphA4 receptors lead to bound and un-bound EphA4 receptors \citep{hornberger_modulation_1999}. The number of cis-bound EphA4 receptors is assumed to be $r_{\!\scriptscriptstyle A4}^{\mathrm{cis}} = w_{\!\scriptscriptstyle A4} \times r_{\!\scriptscriptstyle A4} \times l_0$ where $w_{\!\scriptscriptstyle A4} = 0.153$ is a binding weight modelling the probability that an EphA4 receptor will bind to a nearby retinal ephrinA ligand. Remaining, free EphA4 receptors are $r_{\!\scriptscriptstyle A4}^{\mathrm{free}} = r_{\!\scriptscriptstyle A4} - r_{\!\scriptscriptstyle A4}^{\mathrm{cis}}$ (Fig.\,\ref{f:clustermech}C).

It is known that EphA receptors form clusters whose presence is necessary for ligand attachments to induce signal transmission and that cluster size affects signal size \citep{nikolov_ephephrin_2013}. It is also known that that some receptor sub-types may compete for ligand attachment \citep{fiore_regulation_2019}.
Additionally, EphA4 receptors make `side attachments' to clusters of EphA receptors, possibly regulating the cluster size \citep{nikolov_ephephrin_2013}. 
%
We focused on the idea that signal strength is related to receptor cluster size and defined $c_i$, the effective cluster size for receptor $i$. 
We set $c_i=1$ for $i=1$, $2$ or $3$ and investigated a simple relationship between EphA cluster size, $c_0$, and $r_{\!\scriptscriptstyle A4}$, the density of EphA4 receptors:
\begin{equation}
    c_0 = 1/r_{\!\scriptscriptstyle A4}
\end{equation}
%
Eq.\,\ref{e:G} was modified to include the cluster size:
%
\begin{equation}\label{e:Gcs}
\mathbf{G} = m_{\!_G}\,\sum_i^N F_i\,c_i\,r_{i,b} \nabla L_i(\mathbf{x})
\end{equation}
